---
title: "Imogen's Metric Analysis"
author: "Imogen Meers"
date: "2/9/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set Up

## Packages

Let's load in the packages we will need for this analysis:

```{r warning=FALSE}
# install.packages("Hmisc)
# install.packages("matrixStats")
# install.packages("reshape")
# install.packages("ggplot2")
# install.packages("rlist")
# install.packages("xgboost")

library(Hmisc) # Load HMisc
library(matrixStats) # Load matrix stats
library(reshape) # Load reshape
library(ggplot2) # Load ggplot2
library(rlist) # Load rlist
library(xgboost) # Load XGBoost
library(readxl)
library(dplyr)
```



For this analysis, we are going to be looking at personal metric data. You should be able to load in your own data in place of the provided data and run this for your own metrics:


```{r}
# Load in data
load("metric_data_2020.rda")

metric_data = read_excel("./Personal Metric Imogen.xlsx")
metric_data$`Screen_Time` = metric_data$`Screen_Time` + (metric_data$`...8`/60)
metric_data$`Sleep_Hours` = metric_data$`Sleep_Hours` + (metric_data$`...6`/60)


metric_data <- metric_data %>% select(-`...6`, -`...8`) %>% mutate(Weekend = ifelse(Day == 'Saturday' | Day == 'Sunday', 1, 0)) 

```


# Sleep Analysis

Sleep is a very important metric but often a single night of bad sleep is not going to have a massively detrimental effect. However, prolonged sleep deprivation is likely to negative consequences. To evaluate this, we are going to create a few derived variables to look at lagged sleep. 

We want to look at this over different time periods so let's create a function to do this. 

Let's start by writing the code to do it for a single metric, sleep.

```{r}
# Set parameter values
lagged_days <- 7
metric <- "Sleep_Hours"
include_current <- TRUE

# Extract vector to lag
met_vec <- unlist(metric_data[,metric])

if(include_current){
    # Create empty data frame to store results
    res_data <- matrix(NA, nrow = nrow(metric_data), ncol = lagged_days + 1)
    
    # Add current column to res_data
    res_data[,lagged_days + 1] <- met_vec
  } else{
     # Create empty data frame to store results
    res_data <- matrix(NA, nrow = nrow(metric_data), ncol = lagged_days)
  
}
  



# Loop through days and calculate lagged values
for(i in 1:lagged_days){
  # Store lagged values
  res_data[,i] <- Lag(x = met_vec, shift = i)
}

# Create result vectors 
mean_val <- sum_val <- max_val <- min_val <- missing_vals <- rep(NA, nrow(metric_data))

# Calculate results for each day:
mean_val <- rowMeans(res_data, na.rm = TRUE)
sum_val <- rowSums(res_data, na.rm = TRUE)
max_val <- rowMaxs(res_data, na.rm = TRUE)
min_val <- rowMins(res_data, na.rm = TRUE)
missing_vals <- rowSums(is.na(res_data))
# Combine results vector
results <- cbind.data.frame(mean_val, sum_val, max_val, min_val, missing_vals)

  # Handle the infinite values for max and min calculations
  results$max_val[is.infinite(results$max_val)] <- NA
  results$min_val[is.infinite(results$min_val)] <- NA

```

Let's view what this looks like:

```{r}
head(results) # View first rows
tail(results) # View last rows
```

This looks good, let's make it into a function:


```{r}
lag_metric_calculator <- function(metric_data, lagged_days = 3,
                                  metric = "Sleep_Hours",
                                  include_current = TRUE){
  #'
  #' This function carries out calculations on lagged values of a given
  #' metric
  #' 
  #' @param metric_data The data frame containing the metrics of interest
  #' @param lagged_days How many lagged days to use in the calculations
  #' @param metric The metric to create the calculations for 
  #' @param include_current Should the current day be included in the 
  #' calculation
  #'
  #' @return This function returns a data frame containing the calculated mean,
  #' sum, max, and min for the set time period (lagged_days). It also calculates
  #' how many values were missing in that time period. 
  #'
  
  # Extract vector to lag
  met_vec <- unlist(metric_data[,metric])
  
  if(include_current){
    # Create empty data frame to store results
    res_data <- matrix(NA, nrow = nrow(metric_data), ncol = lagged_days + 1)
    
    # Add current column to res_data
    res_data[,lagged_days + 1] <- met_vec
  } else{
     # Create empty data frame to store results
    res_data <- matrix(NA, nrow = nrow(metric_data), ncol = lagged_days)
  
  }
  
  
  # Loop through days and calculate lagged values
  for(i in 1:lagged_days){
    # Store lagged values
    res_data[,i] <- Lag(x = met_vec, shift = i)
  }
  
  # Calculate results for each day:
  mean_val <- rowMeans(res_data, na.rm = TRUE) # Calculate mean value
  sum_val <- rowSums(res_data, na.rm = TRUE) # Calculate sum value
  max_val <- rowMaxs(res_data, na.rm = TRUE) # Calculate max value
  min_val <- rowMins(res_data, na.rm = TRUE) # Calculate min value
  missing_vals <- rowSums(is.na(res_data)) # Calculate missing values
  
  # Combine results vector
  results <- cbind.data.frame(mean_val, sum_val, max_val, min_val, missing_vals)
  
  # Handle the infinite values for max and min calculations
  results$max_val[is.infinite(results$max_val)] <- NA
  results$min_val[is.infinite(results$min_val)] <- NA
  
  # Return results
  return(results)
}

```

Now let's calculate some summary stats for different windows:

```{r}
lag_2_res <- lag_metric_calculator(metric_data, lagged_days = 1,
                                  metric = "Sleep_Hours",
                                  include_current = TRUE)
lag_3_res <- lag_metric_calculator(metric_data, lagged_days = 2,
                                  metric = "Sleep_Hours",
                                  include_current = TRUE)
lag_7_res <- lag_metric_calculator(metric_data, lagged_days = 6,
                                  metric = "Sleep_Hours",
                                  include_current = TRUE)
```


Let's look at the relationship between these values and some other metrics:

First we can look at productivity:

```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Productivity"

# Add metric to lagged results
analysis_2_res <- cbind.data.frame(lag_2_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_2_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_1 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (2 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Wrap facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_1
```


```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Productivity"

# Add metric to lagged results
analysis_3_res <- cbind.data.frame(lag_3_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_3_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_2 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Add points to plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (3 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Add facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_2
```


This is the most interesting Sleep vs Productivity plot to me as it highlight well the times at which I am most productive.

For min and mean val over the last 3 days, I am most productive when this value is in the middle (maybe slightly towards the higher end). It drops off when I sleep too much and too little. For sum of over the last three days, you can also see a peak at around 24 hours, but a drop off if it is more or less. These results are unsurprising as low values of sleep will affect my productivity if I am not able to function as well or maybe I am taking more naps because I am not sleeping well at night. But consistent high values over 3 days also have a negative impact on productivity, which can be explained by the fact that I am probably so exhausted that I cannot be productive.


```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Productivity"

# Add metric to lagged results
analysis_7_res <- cbind.data.frame(lag_7_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_7_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_3 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Add points to plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (7 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Add facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_3
```

Next let's look at wake up:

```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Wake_Up"

# Add metric to lagged results
analysis_2_res <- cbind.data.frame(lag_2_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_2_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_4 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (2 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Wrap facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_4
```

```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Wake_Up"

# Add metric to lagged results
analysis_3_res <- cbind.data.frame(lag_3_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_3_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_5 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (3 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Wrap facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_5
```

```{r warning=FALSE}
# Set metric to look at 
metric_1 <- "Wake_Up"

# Add metric to lagged results
analysis_7_res <- cbind.data.frame(lag_7_res, metric_data[,metric_1])

# Melt data to long form
plot_res <- melt(analysis_7_res, id.vars =metric_1)
names(plot_res)[1] <- "metric_1"

# Create plot
g_6 <-  ggplot(plot_res, # Set data
               aes(x = value, y = metric_1, color = variable)) + # Set aesthetics
  geom_point(alpha = 0.3) + # Set geom point for scatter plot
  geom_smooth() + # Add smoothing line
  labs(y = paste(metric_1), # Add labels
       title = paste(metric_1, "v", "Sleep Hours (7 day lag)")) +
  facet_wrap(~variable, scales = "free") + # Wrap facets
   theme(axis.line = element_line(colour = "black"), # Set axis line as black
        panel.grid.major = element_blank(), # Remove grid
        panel.grid.minor = element_blank(), # Remove grid
        panel.border = element_blank(), # Remove grid
        panel.background = element_blank())  # Remove grid 
# Generate plot
g_6
```


## Adding Modelling

Let's see what predictive power is present here for some metrics:


```{r}
# Create vector of metrics
metrics <- c("Wake_Up", "Screen_Time", "Pickups",
             "Productivity", "Sleep_Hours", "Weekend")
# Create vector of lags
lags <- c(1,3,7)


# Create list to store results
res_list <- vector(mode = "list", length = length(metrics) * length(lags))

# For each lag
for(i in 1:length(lags)){
  # For each metric
  for(j in 1:length(metrics)){
    if(metrics[j] == "Sleep_Hours"){
          # Calculate lagged summary stats
    temp <- lag_metric_calculator(metric_data, lagged_days = lags[i],
                                  metric = metrics[j],
                                  include_current = TRUE)
    } else {
              # Calculate lagged summary stats
    temp <- lag_metric_calculator(metric_data, lagged_days = lags[i],
                                  metric = metrics[j],
                                  include_current = FALSE)
    }

    # Set names
    names(temp) <- paste(names(temp), metrics[j], lags[i], sep ="_")
    
    # Store Results
    res_list[[(j + ((i-1) * length(metrics)))]] <- temp
  }
}

```

Let's join the results together
```{r}
m_dat <- list.cbind(res_list)
```

Let's choose a variable to use as the response:

```{r}
# Assign response variable
response <- metric_data$Productivity
```

Let's drop rows where we have missing values for the response variable
```{r}
# Drop missing response rows
m_dat <- m_dat[!is.na(response),]
response <- response[!is.na(response)]
```


Now let's set our data up for modeling:

```{r}
dtrain <- xgb.DMatrix(data = as.matrix(m_dat), label = response)
```

Let's run a model:

```{r}
fit_1 <- xgboost(dtrain,  # Set dataset to use
                 nrounds = 100, # Set number of rounds
               
               verbose = 1, # 1 - Prints out fit
                print_every_n = 20 # Prints out result every 20th iteration
)
```

Let's load in the SHAP functions:

```{r}
# Load SHAP functions
source("a_insights_shap_functions.r")
```

Now let's generate some SHAP charts:

```{r}
# Calculate SHAP importance
shap_result <- shap.score.rank(xgb_model = fit_1, 
                X_train =as.matrix(m_dat),
                shap_approx = F)
# Calculate data for SHAP plot
shap_long = shap.prep(shap = shap_result,
                           X_train = as.matrix(m_dat), 
                           top_n = 10)

# Generate SHAP plot
plot.shap.summary(data_long = shap_long)
```

From this SHAP plot, we can see the Sleep Hours is generally the most important factor regarding Productivity for me. My top factor is my minimum value of sleep over the last 3 days and when this value is high it has a positive impact of productive. This probably means that I am greatly impacted by one night of bad sleep rather than multiple days of ok sleep. 

Interestingly, my second strongest factor is the sum of my phone pickups across the last 7 days. This is more spread out than the min sleep, but seems to show higher values have a positive impact on productivity. This could be because if I am picking my phone up more in the previous days maybe I am doing less work so I feel the need to be more productive today!

All of my values in this SHAP plot have a positive correlation with productivity, meaning the higher their values, the higher my productivity and most of them have their positive values with high feature value, meaning the high values of those metrics affect the model most. However, there are some metrics of sleep, such as mean sleep in the last 3 days and yesterday, and max sleep yesterday, that have low values with high feature value. This means that my recent sleep habits can also have a negative effect on my productivity if they are particularly bad as well as having a positive effect if they are particularly good.

I added a weekend value to this SHAP plot to see if there are any particular patterns with my productivity on the weekend vs weekdays that could skew my other factors (e.g. I get lots of sleep on Saturday, but I am not productive on Sunday) but it proved to not be a very important factor.





































































